Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-08-07 pot/LaTeX/jsm2014.tex
small changes and tweak to mapping file
jsm updates

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
    \end{itemize}
    \item Spatial extreme borrows strength across space to estimate return levels and make predictions at unknown locations..
    \item In geostatistical analysis, kriging uses spatial correlation for prediction.
    \item Want to explore similar methods for extremes.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard analysis - Peaks over threshold}
  \begin{itemize}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and fit data above the threshold using the generalized Pareto distribution
      \item Temporal dependence may be an issue between observations (e.g. flood levels don't dissipate overnight)
    \item Generalized Pareto distribution (GPD) is used for the exceedances.
  \end{itemize}
\end{frame}

@@

\begin{frame}{$\chi$ coefficient}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item The $\chi$ coefficient is a measure of extremal dependence
   \item Specifically, we focus on $\chi(\bh)$ for the upper tail given by
    \begin{align*}
      \chi(\bh) = \lim_{c \rightarrow \infty}\Pr(Y(\bs) > c \mid Y(\bs + \bh) > c)
    \end{align*}
    \item If $ \chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
    \item We expect $\lim_{\bh \rightarrow \infty}\chi(h)=0$.
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function
$\rho(\bh)$.
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(c) = 0$ regardless of the strength of the correlation in the bulk of the distribution.
    \item Tail is not flexible (Gaussian is light tailed)
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by skewness $\alpha$ and degrees of freedom $2a$
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(c) > 0$ Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(\bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$ 
   \item This occurs because all observations (near and far) share the same $z_t$ and $\sigma_t^2$.
    \item We deal with this through a daily random partition (similar to Huser and Davison).
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $w_{tk} \sim$ Uniform that define a random daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - w_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MV skew-t distribution as before.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Simulated $\chi$ plots}
  \centering
  \begin{figure}
  \includegraphics[width=1\linewidth]{./plots/chi-plots.pdf}\\[-0.25in]
  \includegraphics[width=0.2\linewidth]{./plots/chi-legend.pdf}
  \caption{$\chi$ plots from simulated datasets}
  \end{figure}
\end{frame}

@@

\begin{frame}{Spatiotemporal Model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item We can account for time in one of two ways
    \begin{itemize}
    	\item The mean: e.g. AR(1)
	\item Three dimensional covariance model (e.g. Huser and Davison, 2014) 
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute missing observations and censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

\begin{frame}{Data analysis}
	\centering
	\begin{figure}
    \includegraphics[width=0.8\linewidth]{./plots/ozone_station.pdf}
    \caption{Ozone monitoring station locations}
    \end{figure}

@@

\begin{frame}{Exploratory data analysis}
	\centering
  \begin{figure}
    \includegraphics[width=1\linewidth]{./plots/chi-plot-ozone-res.pdf}
    \caption{$\chi$-plot for residuals of ozone sample quantiles}
  \end{figure}
\end{frame}

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods incorporating
    \begin{itemize}
      \item Gaussian vs $t$ vs skew-$t$ marginal distribution
      \item $K=1$ partition vs $K=3$ partitions
      \item No thresholding vs thresholding at $T=0.90$ sample quantile
    \end{itemize}
    \item All methods use a \Matern or exponential covariance ($\nu = 0.5$)
    \item Compare quantile and Brier scores using 5-fold cross validation (Gneiting and Raftery, 2007)
    \item Mean function modeled using a first-order spatial trend
  \end{itemize}
\end{frame}

@@

\begin{frame}{Five-fold cross-validation results}
  \begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{l|c|r|rrrrr}
          \multicolumn{3}{c}{\ } & \multicolumn{5}{c}{Quantile}\\
           \hline
  Marginal & $K$ & $T$  & 0.900 & 0.950 & 0.990 & 0.995 & 0.999\\
  \hline
Gaussian & 1 & 0 & 3.807 & 2.981 & 1.797 & 1.493 & 1.044\\
$t$ & 1 & 0 & 2.622 & 1.584 & 0.436 & 0.244 & 0.063 \\
$t$ & 3 & 0 & 2.643 & 1.601 & 0.455 & 0.265 & 0.085 \\
skew-$t$ & 1 & 0 & 2.766 & 1.683 & 0.467 & 0.257 & 0.063 \\
$t$ & 1 & 0.9 & 4.514 & 2.422 & 0.577 & 0.318 & 0.086 \\
skew-$t$ & 1 & 0.9 & 4.396 & 2.388 & 0.573 & 0.311 & 0.079 \\
\hline
    \end{tabular}
    \caption{Selected five-fold cross-validation results}
  \end{table}
  \begin{itemize}
  	\item Results for 
  	\item Brier score results are similar.
  \end{itemize}
\end{frame}

\begin{frame}{Simulation study}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 6 different data settings:
    \begin{itemize}
    	\item Gaussian vs $t$ vs skew-$t$ marginal distribution
        \item $K=1$ partition vs $K=5$ partitions
    \end{itemize}
    \item Results are similar to the results from the data analysis
    \item Biggest gains come from thresholding.
    \item Using skew models give additional gain, but small relative to gain for thresholding.
  \end{itemize}
\end{frame}

\begin{frame}{Future work}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Comparison with extreme value analysis methods
    \item Reporting of compliance with EPA ozone standards
    \item Including time in the model via standard spatiotemporal Gaussian models.  
  \end{itemize}
\end{frame}

>>>>

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
      \item e.g. Probability of exceeding a certain threshold level
    \end{itemize}
    \item Spatial methods borrow information across space to estimate spatial correlation and make predictions by Kriging at unknown locations.
    \item Want to explore similar methods for extremes.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard analysis - Peaks over threshold}
  \begin{itemize}  \setlength{\itemsep}{0.5em}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and use the Generalized Pareto distribution (GPD) to model the exceedances.
      \item Temporal dependence may be an issue between observations (e.g. flood levels don't dissipate overnight)
  \end{itemize}
\end{frame}

@@

\begin{frame}{$\chi$ coefficient}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item The $\chi$ coefficient is a measure of extremal dependence
   \item Specifically, we focus on $\chi(\bh)$ for the upper tail given by
    \begin{align*}
      \chi(\bh) = \lim_{c \rightarrow \infty}\Pr(Y(\bs) > c \mid Y(\bs + \bh) > c)
    \end{align*}
    \item If $ \chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
    \item We expect $\lim_{\bh \rightarrow \infty}\chi(\bh)=0$.
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function
$\rho(\bh)$.
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(\bh) = 0$ regardless of the strength of the correlation in the bulk of the distribution.
    \item Tail is not flexible (Gaussian is light tailed)
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by skewness $\alpha$ and degrees of freedom $2a$
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(\bh) > 0$ (Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(\bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$
   \item This occurs because all observations (near and far) share the same $z_t$ and $\sigma_t^2$.
    \item We deal with this through a daily random partition (similar to Huser and Davison).
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $\bw_{tk} \sim$ Uniform that define a random daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - \bw_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MV skew-t distribution as before.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Simulated $\widehat{\chi}(\bh)$ plots}
  \centering
  \begin{figure}
  \includegraphics[width=1\linewidth]{./plots/chi-plots.pdf}\\[-0.25in]
  \includegraphics[width=0.2\linewidth]{./plots/chi-legend.pdf}
  \caption{$\chi$ plots from simulated datasets}
  \end{figure}
\end{frame}

\begin{frame}{Sample simulated datasets}
  \centering
  \begin{figure}
  \includegraphics[width=1\linewidth]{./plots/gauss-vs-t3.pdf}
  \caption{Gaussian and $t$ with 3 partitions}
  \end{figure}
\end{frame}

@@

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

\begin{frame}{Data analysis}
\begin{columns}[c]
\column{.45 \linewidth}
	\begin{itemize} \setlength{\itemsep}{0.5em}
	\item Data analysis uses
	\begin{itemize}
		\item max 8-hour ozone measurements
		\item 85 sites
		\item 92 days
	\end{itemize}
	\end{itemize}

	\column{.55\linewidth}
	\begin{figure}
    \includegraphics[width=1\linewidth]{./plots/ozone_station.pdf}
    \caption{Ozone monitoring station locations}
    \end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{Exploratory data analysis}
	\centering
  \begin{figure}
    \includegraphics[width=1\linewidth]{./plots/chi-plot-ozone-res.pdf}
    \caption{$\widehat{\chi}$-plot for sample quantiles of ozone observations}
  \end{figure}
\end{frame}

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods incorporating
    \begin{itemize}
      \item Gaussian vs $t$ vs skew-$t$ marginal distribution
      \item $K=1$ partition vs $K=3$ partitions
      \item No thresholding vs thresholding at $T=0.90$ sample quantile
    \end{itemize}
    \item All methods use a \Matern or exponential covariance ($\nu = 0.5$)
    \item Compare quantile and Brier scores using 5-fold cross validation (Gneiting and Raftery, 2007)
    \item Mean function modeled as
    \begin{align*}
    	\beta_0 + \beta_1 \cdot \text{lat} + \beta_2 \cdot \text{long} +  \beta_3 \cdot \text{lat}^2 + \beta_4 \cdot \text{long}^2 + \beta_5 \cdot \text{lat} \cdot \text{long} 
    \end{align*}
   \end{itemize}
\end{frame}

@@

\begin{frame}{Five-fold cross-validation results}
  \begin{table}[htbp]
    \small
    \centering
    \begin{tabular}{l|c|r|rrrrr}
          \multicolumn{3}{c}{\ } & \multicolumn{5}{c}{$\tau$}\\
           \hline
  Marginal & $K$ & $T$  & 0.950 & 0.980 & 0.990 & 0.995 & 0.999\\
  \hline
Gaussian 	& 1 & 0 	& 39.820 & 17.539 & 9.167 & 4.720 & 1.057\\
$t$		& 1 & 0 	& {\bf 31.008} & {\bf 13.898} & 7.229 & {\bf 3.405} & 0.879\\
$t$		& 3 & 0	& 31.213 & 13.920 & {\bf 7.218} & 3.498 & 0.918\\
$t$		& 1 & 0.9	& 32.221 & 14.519 & 7.549 & 3.604 & 0.896\\
$t$		& 3 & 0.9 	& 38.842 & 16.781 & 8.434 & 4.180 & 1.020\\
skew-$t$	& 1 & 0	& 31.845 & 14.542 & 7.533 & 3.645 & {\bf 0.844}\\
skew-$t$	& 1 & 0.9	& 32.132 & 14.296 & 7.484 & 3.497 & 0.890\\
skew-$t$	& 3 & 0	& 33.653 & 15.453 & 8.119 & 4.338 & 1.188\\
skew-$t$	& 3 & 0.9 	& 32.157 & 14.727 & 7.794 & 3.825 & 0.917\\
\hline
    \end{tabular}
    \caption{Brier score for predicting exceedance of $c = \hat{q}(\tau)$ from five-fold cross-validation ($\times 1000$)}
  \end{table} \vspace{-1.5em}
  \begin{itemize}
  	\item Quantile score results are similar.
  \end{itemize}
\end{frame}

\begin{frame}{Predicted 95th quantile}
\centering
\begin{figure}
    \includegraphics[width=.5\linewidth]{./plots/quantile-95-gau.pdf}
    \includegraphics[width=.5\linewidth]{./plots/quantile-95-t19.pdf}
    \caption{Predicted 95th quantile using Gaussian and $t$ thresholded at $T=0.9$}
\end{figure}
\end{frame}

\begin{frame}{Predicted 99th quantile}
\centering
\begin{figure}
    \includegraphics[width=.5\linewidth]{./plots/quantile-99-gau.pdf}
    \includegraphics[width=.5\linewidth]{./plots/quantile-99-t19.pdf}
    \caption{Predicted 95th quantile using Gaussian and $t$ thresholded at $T=0.9$}
\end{figure}
\end{frame}

\begin{frame}{Probability of exceedance}
\centering
\begin{figure}
    \includegraphics[width=.5\linewidth]{./plots/p-exceed-std-gau.pdf}
    \includegraphics[width=.5\linewidth]{./plots/p-exceed-std-t19.pdf}
    \caption{Probability of exceeding the 75 ppb ozone standard using Gaussian and $t$ thresholded at $T=0.9$}
\end{figure}
\end{frame}

\begin{frame}{Simulation study}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 6 different data settings:
    \begin{itemize}
    	\item Gaussian vs $t$ vs skew-$t$ marginal distribution
        \item $K=1$ partition vs $K=5$ partitions
    \end{itemize}
    \item Preliminary results suggest that there may be a modest improvement in quantile scores from thresholding when the data come from a $t$ distribution with $K=5$ partitions.
  \end{itemize}
\end{frame}

\begin{frame}{Future Work}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Different ways to incorporate the temporal dependence
    \begin{itemize}
    	\item Three dimensional covariance model for $v_t(\bs)$ (e.g. Huser and Davison, 2014)
    	\item Use a temporal structure for $z_t(\bs)$:
	\begin{itemize}
		\item AR(1)
		\item Moving average 
		\item Association between $\bw_{t, k}$ and $\bw_{t+1, k}$
	\end{itemize}
    \end{itemize}
     \item Comparison with extreme value analysis methods
  \end{itemize}
\end{frame}
